Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2023-09-15 paper-new/paper.tex
some writing

\title{Denotational Semantics for Gradual Typing in Synthetic Guarded Domain Theory}
\author{Eric Giovannini}
\author{Max S. New}

\begin{abstract}
    We develop a denotational semantics for a simple gradually typed language
    that is adequate and proves the graduality theorem.
    %
    The denotational semantics is constructed using \emph{synthetic
    guarded domain theory} working in a type theory with a later
    modality and clock quantification.
    %
    This provides a remarkably simple presentation of the semantics,
    where gradual types are interpreted as ordinary types in our ambient
    type theory equipped with an ordinary preorder structure to model
    the error ordering.
    %
    This avoids the complexities of classical domain-theoretic models
    (New and Licata) or logical relations models using explicit
    step-indexing (New and Ahmed).
    %
    In particular, we avoid a major technical complexity of New and
    Ahmed that requires two logical relations to prove the graduality
    theorem.

    By working synthetically we can treat the domains in which gradual
    types are interpreted as if they were ordinary sets. This allows us
    to give a ``na\"ive'' presentation of gradual typing where each
    gradual type is modeled as a well-behaved subset of the universal
    domain used to model the dynamic type, and type precision is modeled
    as simply a subset relation.
    %
  \end{abstract}

@@

>>>>

\title{Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory}
\author{Eric Giovannini}
\author{Max S. New}

\begin{abstract}
  Gradually typed programming languages, which allow for soundly
  mixing static and dynamically typed programming styles, present a
  strong challenge for metatheorists. Even the simplest sound
  gradually typed languages feature at least recursion and errors,
  with realistic languages featuring furthermore runtime allocation of
  memory locations and dynamic type tags. Further, the desired
  metatheoretic properties of gradually typed languages have become
  increasingly sophisticated: validity of typed based equational
  reasoning as well as the relational property known as the gradual
  guarantee or graduality. Many recent works have tackled verifying
  these properties, but the resulting mathematical developments are
  highly repetitive and tedious, with few reusable theorems persisting
  across different developments.

  In this work, we present a new denotational account of gradual
  typing semantics developed using guarded domain theory. Guarded
  domain theory combines the expressive power of step-indexed logical
  relations for modeling recursive features with the modularity and
  reusability of denotational semantics. Further, recent extensions to
  cubical Agda mean that synthetic guarded domain theory is readily
  mechanized in a proof assistant. We demonstrate the feasibility of
  this approach with a model of gradually typed lambda calculus and
  prove the validity of beta-eta equality and the graduality theorem
  for the denotational model. This model should provide the basis for
  a reusable mathematical theory of gradually typed program semantics.
%%     We develop a denotational semantics for a simple gradually typed language
%%     that is adequate and proves the graduality theorem.
%%     %
%%     The denotational semantics is constructed using \emph{synthetic
%%     guarded domain theory} working in a type theory with a later
%%     modality and clock quantification.
%%     %
%%     This provides a remarkably simple presentation of the semantics,
%%     where gradual types are interpreted as ordinary types in our ambient
%%     type theory equipped with an ordinary preorder structure to model
%%     the error ordering.
%%     %
%%     This avoids the complexities of classical domain-theoretic models
%%     (New and Licata) or logical relations models using explicit
%%     step-indexing (New and Ahmed).
%%     %
%%     In particular, we avoid a major technical complexity of New and
%%     Ahmed that requires two logical relations to prove the graduality
%%     theorem.

%%     By working synthetically we can treat the domains in which gradual
%%     types are interpreted as if they were ordinary sets. This allows us
%%     to give a ``na\"ive'' presentation of gradual typing where each
%%     gradual type is modeled as a well-behaved subset of the universal
%%     domain used to model the dynamic type, and type precision is modeled
%%     as simply a subset relation.
%%     %
  \end{abstract}

@@

% Outline

% 1. Intro: What do we want out of gradually typed languages and why
%    is it hard to prove? Explanation: Gradual Typing inherently
%    involves recursive types, multiple effects, relational properties.
%    We argue that the increasing complexity of the metatheory of
%    gradual typing makes it a good candidate for

% 2. Extensional Dream Semantics: double categorical

% 3. The Problem with Step-indexing

% 4. A Compromise

% 5. Formalization in Guarded Cubical Agda

### 2023-09-15 paper-new/semantics.tex
some writing

>>>>

\section{Denotational Semantics}

First, we define a denotational semantics of types and terms of the
cast calculus by giving a standard monadic denotational semantics in
the cartesian closed category of preorders and monotone functions,
extended to model the primitives of gradual typing: the dynamic type,
errors and type casts. The most interesting part of this semantics is
the construction of the monad and the dynamic type.

### 2023-09-15 paper-new/technical-background.tex
some writing

>>>>

\subsection{Operational Reduction Proofs}

\subsection{Classical Domain Models}

New and Licata \cite{newlicata} developed an axiomatic account of the
graduality relation on cast calculus terms and gave a denotational
model of this calculus using classical domain theory based on
$\omega$-CPOs. This semantics has scaled up to an analysis of a
dependently typed gradual calculus in \cite{asdf}. This meets our
criterion of being a reusable mathematical theory, as general semantic
theorems about gradual domains can be developed independent of any
particular syntax and then reused in many different denotational
models. However, it is widely believed that such classical domain
theoretic techniques cannot be extended to model higher-order store, a
standard feature of realistic gradually typed languages such as Typed
Racket. Thus if we want a reusable mathematical theory of gradual
typing that can scale to realistic programming languages, we need to
look elsewhere to so-called ``step-indexed'' techniques.

A series of works \cite{newahmed,newlicataahmed,newjamnerahmed}
developed step-indexed logical relations models of gradually typed
languages based on operational semantics. Unlike, classical domain
theory, such step-indexed techniques are capable of modeling
higher-order store and runtime-extensible dynamic types
\cite{amalsthesis,nonpmetricparamorsomething,newjamnerahmed}. However
their proof developments are highly repetitive and technical, with
each development formulating a logical relation from first-principles
and proving many of the same tedious lemmas without reusable
mathematical abstractions. Our goal in the current work is to extract
these reusable mathematical principles from these tedious models to
make formalization of realistic gradual languages tractible.
